---
title: "Assignment 04: Tidy data and joins"
author: "Diana Lin"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set Up

Load required packages:

```{r}
suppressMessages(library(tidyverse))
suppressMessages(library(gapminder))
suppressMessages(library(DT))
```

## Exercise 1: Univariate Data Reshaping

Choose **EITHER** “Univariate Option 1” or “Univariate Option 2”. Both of these problems have three components:

1. Putting data in wider format;
1. Producing a plot of the wide data;
1. Re-lengthening the wider data. 

You are expected to use `pivot_wider()` and `pivot_longer()` for reshaping, and ggplot2 for the plot.

Regarding the plot:

* Don’t worry about plotting all of the data from the wide tibble.
* The plot should be something that someone might actually want to produce in an analysis.

### Univariate Option 1
#### 1. Make a tibble with one row per year, and columns for life expectancy for two or more countries.

```{r}
widened <- gapminder %>% 
  pivot_wider(id_cols = year, names_from = "country", values_from = lifeExp)
```

* Each row in this table corresponds to a single year, and there is a column for each of the 142 countries containing the life expectancy in years.
* There is a total of only 12 rows in this `tibble`, as there are only 12 years in the gapminder data set.

#####  __NOTE__: this datatable can be scrolled **horizontally** to reveal _more_ columns!

```{r, echo = FALSE}
widened %>%
  datatable(rownames = FALSE,
            options = list(scrollX = TRUE, fixedColumns = TRUE))
```

#### 2. Take advantage of this new data shape to scatterplot life expectancy for one country against that of another.

```{r}
widened_plot <- widened %>%
  ggplot(aes(Canada, China)) +
  geom_point(aes(color = as.factor(year))) +
  ggtitle("Life Expectancy from 1952 to 2007") +
  ylab("Life Expectancy in China (years)") +
  xlab("Life Expectancy in Canada (years)") +
  scale_color_discrete("Year") +
  theme(text = element_text(size=18))
```

* China consistently has a lower life expectancy than Canada in every year.
* Both countries have a steadily increasing life expectancy overall, except for China in 1962, where the China's life expectancy dropped substantially.

```{r, echo = FALSE}
print(widened_plot)
```



#### 3. Re-lengthen the data.

```{r}
lengthened <- widened %>%
  pivot_longer(cols = -year, names_to = "country", values_to = "lifeExp")
```

* The relengthened `tibble` will _only_ contain `year`, `country`, `lifeExp` (since the other columns `continent`, `gdpPercap` and `pop` were dropped during the original `pivot_wider()`), with each combination of `country` and `year` in each row.

```{r, echo = FALSE}

datatable(lengthened, rownames = FALSE)
```

## Exercise 2: Multivariate Data Reshaping
Choose **EITHER** “Multivariate Option 1” or “Multivariate Option 2”. All of these problems have two components:

1. Putting data in wider format;
1. Re-lengthening the data.

Don’t worry about producing a plot here. You are expected to use `pivot_wider()` and `pivot_longer()` for reshaping.

### Multivariate Option 1

#### 1. Make a tibble with one row per year, and columns for life expectancy and GDP per capita (or two other numeric variables) for two or more countries.

```{r}
widened <- gapminder %>%
  pivot_wider(id_cols  = year, 
              names_from = country,
              names_sep = "_", 
              values_from = c("lifeExp", "gdpPercap"))
```

* This widened `tibble` contains a column for `lifeExp` and `gdpPercap` of all 142 countries for each row, meaning a total of 285 columns.
* This `tibble` has 12 rows, as that is the number of unique years in the `gapminder` dataset.

##### __NOTE__: this datatable can be scrolled **horizontally** to reveal _more_ columns!

```{r, echo = FALSE}
datatable(widened, rownames = FALSE, 
          options = list(scrollX = TRUE, fixedColumns = TRUE))
```

#### 2. Re-lengthen the data.

```{r}
lengthened <- widened %>%
  pivot_longer(cols = -year,
               names_to = c(".value", "country"),
               names_sep = "_")
```

* The relengthened `tibble` below will _only_ contain `year`, `country`, `lifeExp` and `gdpPercap` (since the other columns `continent` and `pop` were dropped during the original `pivot_wider()`), with each combination of `country` and `year` in each row.

```{r, echo = FALSE}
datatable(lengthened, rownames = FALSE)
```

## Exercise 3: Table Joins

Do **ALL** of the activities in this section.

Read in the made-up wedding guestlist and email addresses using the following lines (go ahead and copy-paste these):

```{r}
guest <- suppressMessages(
  read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
  )

email <- suppressMessages(
  read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")
  )
```

Then, complete the following tasks using the `tidyverse` (`tidyr`, `dplyr`, …). No need to do any pivoting – feel free to leave `guest` in its current format.


### Exercise 3.1

#### For each guest in the guestlist (`guest` tibble), add a column for email address, which can be found in the `email` tibble.

* In order to join these two tibbles, the `email` tibble must be separated by guest as opposed to by party, using `separate_rows()`.
* Then, the guest column in `email` must be renamed to `name`, as it appears in the `guest` tibble, using `rename()`.

```{r}
email_sep <- email %>%
  separate_rows(email, guest, sep = ", ") %>%
  rename(name = guest)
```

* Then, since additional information is to be added to the original tibble, we employ the use of `left_join()`.

```{r}
guest_email <- left_join(guest,
          email_sep,
          by = "name"
)
```

* This datatable contains all guests on the guest list, and an email address if one was provided.

##### __NOTE__: this datatable can be scrolled **horizontally** to reveal _more_ columns!

```{r, echo = FALSE}
datatable(guest_email, rownames = FALSE,
          options = list(scrollX = TRUE, fixedColumns = TRUE))
```

### Exercise 3.2

#### Who do we have emails for, yet are not on the guestlist?

* Using `anti_join()`, guests in `email_sep` who are also in `guest` will be dropped.

```{r}
email_not_guest <- anti_join(email_sep,guest, by = "name")
```

* There are only three people we have emails for that are not on the guest list.
* Since the original `email` tibble was processed to be separated by `name` instead of party, it is possible that other members of the party are on the guest list but these three are not.
* After double checking, it is revealed that these three below belong to two separate parties, and none of the other party members were on the original guest list either.

```{r, echo = FALSE}
datatable(email_not_guest, rownames = FALSE)
```

### Exercise 3.3

#### Make a guestlist that includes everyone we have emails for (in addition to those on the original guestlist).

* The tibble should combine all guests in the `guest` list, and all guests in `email_sep`, by using `full_join()`.
```{r}
all <- full_join(guest, email_sep, by = "name")
```

* The three people with emails but aren't on the guest list have been appended to the end of the `tibble`.
* They have `NA` values for almost every column except for `name` and `email`.

##### __NOTE__: this datatable can be scrolled **horizontally** to reveal _more_ columns!

```{r, echo = FALSE}
datatable(all, rownames = FALSE,
          options = list(scrollX = TRUE, fixedColumns = TRUE))
```

