---
title: "hw02_gapminder-dplyr"
output: html_document
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set Up
Load the necessary packages for this assignment:
```{r load}
library(gapminder)
library(tidyverse)
```

## Exercise 1: Basic `dplyr`
### Exercise 1.1
Use `filter()` to subset the `gapminder` data to three countries of your choice in the 1970s.
```{r}
filter(gapminder, year >= 1970 &
           year < 1980, 
       country == "Canada" | 
           country == "China" | 
           country == "Japan")
```

### Exercise 1.2
Use the pipe operator `%>%` to select "country" and "gdpPercap" from your filtered dataset in 1.1.
```{r}
filter(gapminder, year >= 1970 & 
           year < 1980, 
       country == "Canada" | 
           country == "China" | 
           country == "Japan") %>%
    select("country", "gdpPercap")
```

### Exercise 1.3
Filter gapminder to all entries that have experienced a drop in life expectancy. Be sure to include a new variable that’s the increase in life expectancy in your tibble. Hint: you might find the `lag()` or `diff()` functions useful.



### Exercise 1.4
Filter gapminder so that it shows the max GDP per capita experienced by each country. Hint: you might find the `max()` function useful here.

```{r}
gapminder %>%
    group_by(country) %>% 
    summarize(maxGDP = max(gdpPercap))
```


### Exercise 1.5
Produce a scatterplot of Canada’s life expectancy vs. GDP per capita using `ggplot2`, without defining a new variable. That is, after filtering the gapminder data set, pipe it directly into the `ggplot()` function. Ensure GDP per capita is on a log scale.

```{r}
gapminder %>%
    filter(lifeExp, gdpPercap) %>%
    ggplot(aes(gdpPercap, lifeExp)) +
    geom_point() + 
    scale_x_log10()
```


## Exercise 2: Explore individual variables with `dplyr`

Pick one categorical variable and one quantitative variable to explore. Answer the following questions in whichever way you think is appropriate, using `dplyr`:

#### What are possible values (or range, whichever is appropriate) of each variable?
* Categorical variable: `country`
* Possible values for the categorical variable are the categories itself. The range would be however many categories there are.
* THere are 142 countries in this dataset.
```{r}
gapminder %>%
    select(country) %>%
    unique() %>% 
    nrow()
```
* Quantitative variable: `lifeExp`
* Possible values (or ranges) are only applicable to quantitative variables. In the case of `lifeExp`, a range of 0 to the maximum life expectancy in the dataset (rounded to a whole number) would be an appropriate range.
* The highest life expectancy is 83 years in this dataset.

```{r}
gapminder %>%
    select(lifeExp) %>%
    max() %>%
    round()
```

#### What values are typical? What’s the spread? What’s the distribution? Etc., tailored to the variable at hand. Feel free to use summary stats, tables, figures.

* A typical value for life expectancy is anywhere from 0 to 100 (realistically), but in this dataset the minimum is 23.60 years and the maximum is 82.60 years.
* A typical value for country is any of the 142 countries included in this.

```{r}
summary(gapminder)
```



## Exercise 3: Explore various plot types

Make two plots that have some value to them. That is, plots that someone might actually consider making for an analysis. Just don’t make the same plots we made in class – feel free to use a data set from the `datasets` R package if you wish.

1. A scatterplot of two quantitative variables.
2. One other plot besides a scatterplot.

You don’t have to use all the data in every plot! It’s fine to filter down to one country or a small handful of countries.

## Recycling (Optional)
Evaluate this code and describe the result. Presumably the analyst’s intent was to get the data for Rwanda and Afghanistan. Did they succeed? Why or why not? If not, what is the correct way to do this?

`filter(gapminder, country == c("Rwanda", "Afghanistan"))`

## Tibble display
Present numerical tables in a more attractive form using `knitr::kable()` for small tibbles (say, up to 10 rows), and `DT::datatable()` for larger tibbles.