---
title: "Gapminder Exploration"
output: ioslides_presentation
---

```{r load, echo = FALSE}
library(gapminder)
library(tibble)
```
## Function #1: `mean()`
Calculate the average life expectancy of all countries in `gapminder`:
```{r}
mean(gapminder$lifeExp)
```

Calculate the average life expectancy of Canada:
```{r}
mean(gapminder[which(gapminder$country == "Canada"),]$lifeExp)
```

## Function #1: `mean()` {.smaller}

Calculate the continental average for each numeric data column:
```{r}
header <- c("Life Expectancy", "Population", "GDP per Capita")
temp <- data.frame()
for (continent in unique(gapminder$continent)) {
    current <- c(
        mean(gapminder[gapminder$continent==continent,]$lifeExp, na.rm = TRUE), 
        mean(gapminder[gapminder$continent==continent,]$pop, na.rm = TRUE),
        mean(gapminder[gapminder$continent==continent,]$gdpPercap, na.rm=TRUE))
    temp <- rbind(temp,current)
}
names(temp) <- header
rownames(temp) <- unique(gapminder$continent)

temp
```

## Function #2: `max()` and `which()`
Extract the country (or countries) with the highest life expectancy:

```{r}
gapminder[which(gapminder$lifeExp == max(gapminder$lifeExp)),]
```

## Function #3: `min()` and `which()`
Extract the country (or countries) with the lowest GDP per capita:

```{r}
gapminder[which(gapminder$gdpPercap == min(gapminder$gdpPercap)),]
```

## Function #4: `length()` and `unique()`
Extract the number of unique countries in `gapminder`:
```{r}
length(unique(gapminder$country))
```

## Function #5: `summary()` {.smaller}
Get a summary of `gapminder`:
```{r}
summary(gapminder)
```
